Mining the human microbiome for clinical insight

### 2018-11-27 abstract.tex
Edit abstract + first part of intro

Human-associated microbial communities are essential for health and have been implicated in many diseases.
DNA sequencing technology has enabled the detailed characterization of these communities, leading to a rapid expansion in studies investigating relationships between the human microbiome and disease.
However, identifying clinically-relevant associations from microbiome datasets is complicated by the high dimensional nature of the data and variability of results across studies.
In this thesis, I describe three projects which overcome various analysis challenges to identify clinically-relevant associations between the human microbiome and disease.
In the first project, I present an analysis of lung, stomach, and oropharyngeal microbiomes of pediatric patients with aerodigestive symptoms.
I leverage data collected from multiple sites per patient to identify  clinically-actionable alterations in aerodigestive community relationships in patients with swallowing dysfunction and to discover new characteristics of the human lung and stomach microbiomes.
In the second project, I perform a meta-analysis of case-control gut microbiome datasets across 28 studies and 10 diseases.
By standardizing processing and analysis methods across many datasets, I find consistent disease-specific and shared patterns of associations which can inform therapeutic development and clinical treatment approaches.
In the third project, I describe a framework for performing rational donor selection in fecal microbiota transplant clinical trials.
In this framework, microbial associations identified from clinical and basic research are used to inform the donor selection strategy, increasing the likelihood of a successful clinical trial.
These projects demonstrate a variety of approaches for mining human microbiome data to identify clinically-relevant associations and discover new fundamental properties of human-associated microbial communities.
Together, this work suggests multiple avenues forward for translating findings from microbiome data analyses into clinical impact.

>>>>

The human microbiome is essential for health and implicated in many diseases.
DNA sequencing has enabled the detailed characterization of human-associated microbial communities, leading to a rapid expansion in studies investigating the human microbiome and disease.
However, extracting clinically-relevant associations from microbiome datasets remains challenging because of the high dimensional nature of the data and variability across studies.
In this thesis, I describe three projects which overcome various analysis challenges to extract useful clinical insights from microbiome data.
In the first project, I present an analysis of lung, stomach, and oropharyngeal microbiomes.
I leverage data collected from multiple sites per patient to identify aspiration-associated changes in the relationships between aerodigestive communities which new approaches for treatment.
%These changes suggest new approaches for developing diagnostics and treatments for aspiration-related respiratory complications.
%I leverage data collected from multiple sites per patient to find aspiration-associated changes in the relationships between aerodigestive communities which suggests new targets for treatment and diagnosis.
In the second project, I perform a meta-analysis of case-control gut microbiome datasets with standard data processing and analysis methods.
I find consistent patterns characterizing disease-associated microbiome changes and a set of shared associations which could inform clinical treatment and therapeutic development approaches.
In the third project, I describe a framework for rational donor selection in fecal microbiota transplant clinical trials.
In this framework, knowledge derived from clinical and basic science research is used to inform which donor is selected for fecal transplants, increasing the likelihood of successful trials.
Together, these projects demonstrate a variety of approaches to mine the human microbiome for clinically-relevant insights and suggests multiple avenues forward for translating findings from microbiome data analyses into clinical impact.

### 2018-11-27 chap1.tex
Edit abstract + first part of intro

The human microbiome, made up of the microbes that live in and on our bodies, is essential for health and has been implicated in many diseases.
Almost all human body sites are colonized by microbial communities, ranging from the gut, which is the largest human-associated microbial community, to the lungs, which were for many years considered sterile (citation needed).
These communities perform essential functions for health, including fighting off and preventing infections, interacting with and regulating host immune responses and metabolism, and metabolizing xenobiotics or other compounds which are indigestible by the host.
On the other hand, perturbations in microbial communities have been implicated in many diseases, including metabolic, autoimmune, neurological, and respiratory conditions.
Thus, the potential for microbiome-based therapies to improve human health and address a broad range of diseases has led to a recent expansion in research and clinical studies in this field.
Much of the emerging research has been driven in part by the increasing accessibility of DNA sequencing technology, which can provide a detailed view of the bacteria in these communities without the need for time-consuming and difficult culturing experiments.

However, despite this increase in microbiome research and its potential for clinical impact, identifying clinically-relevant associations from microbiome studies remains challenging.
Microbiome datasets are high-dimensional, with hundreds to thousands of bacterial species per patient sample but frequently sampling only tens to hundreds of patients.
Microbial communities are also highly variable across people, and identifying  individual bacterial biomarkers that can distinguish health and disease across many patients is difficult.
Finally, microbiome datasets provide a window into associations but not causal relationships.
Researchers must perform follow-up mechanistic studies or FMT clinical trials to confirm the clinical relevance of associations.
In this thesis, we present three projects which overcome various data analysis challenges to provide useful clinical insight from mining microbiome datasets.
The following chapters present three unique analyses of microbiome data, each highlighting a different perspective on analyzing microbiome data in a way that moves beyond reporting statistically significant results to identify clinically meaningful associations.

>>>>

The human microbiome, the microbes that live in and on our bodies, is essential for health and has been implicated in many diseases.
Almost all human body sites are colonized by microbes, ranging from the gut, the largest human-associated microbial community, to the lungs, which were for many years considered sterile (citation needed).
These microbial communities perform essential functions for health, including fighting off and preventing infections, interacting with and regulating host immune responses, and metabolizing xenobiotics or other compounds which are indigestible by the host.
Additionally, perturbations in human microbiomes have been implicated in many diseases, including metabolic, autoimmune, neurological, and respiratory conditions.

The potential for microbiome-based therapies to improve human health and address a broad range of diseases has led to a recent expansion of research and clinical studies in this field.
Much of the emerging research has been driven in part by the increasing accessibility of DNA sequencing technology, which can provide a detailed view of the bacteria in these communities without the need for time-consuming and difficult culturing experiments.
However, identifying clinically-relevant associations from microbiome studies remains challenging.
Microbiome datasets are high-dimensional, with hundreds to thousands of bacterial species measured in usually only tens to hundreds of patient samples.
Microbial communities are also highly variable across people, making it more difficult to identify individual bacterial biomarkers that can consistently distinguish health and disease across many different patients.
Finally, microbiome datasets provide a window into associations but not causal relationships, and researchers must perform follow-up mechanistic studies or clinical trials to confirm the clinical relevance of any identified associations.
In this thesis, we present three unique analyses of microbiome data which overcome some of these challenges and which illustrate a variety of approaches for extracting useful clinical insights from mining microbiome data.
Together, the following chapters aim to move the findings from microbiome data analyses beyond statistical significance and toward clinical meaningfulness.

### 2018-11-27 chap5.tex
Start integrating aspiration paper:
Edit abstract + first part of intro

@@

\epigraph{{\selectlanguage{polutonikogreek}κτῆμά τε ἐς αἰεὶ μᾶλλον
ἢ ἀγώνισμα ἐς τὸ παραχρῆμα ἀκούειν ξύγκειται.}}{Thucydides, \textit{History}, 1.22.4}

>>>>

In the first project, I overcome the difficult of identifying consistent biomarkers across highly variable lung and stomach communities by instead looking at the relationships between different aerodigestive body sites.
In the second project, I re-analyze gut microbiome datasets across many diseases and individual studies to identify consistent associations and overcome the difficulty of extracting generalizable knowledge from the many published studies.
In the third project, I present a framework to leverage insights gleaned from data analyses and mechanistic experiments to improve future fecal microbiota transplant clinical trials.

@@
